---
title: "Session 0: Overview"
description: |
  This session provides a quick overview over the course and a short primer to installing and organizing your R environment. 
author:
  - name: Simon J. Brandl, PhD
    url: https://www.fishandfunctions.com/
    affiliation: The University of Texas at Austin
color1: "#0D2A47"
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: true
    number_sections: true
    toc_depth: 4
    code_folding: true
    toc_float: true
    collapsed: false
    smooth_scroll: true
---

## Structure of coding lessons

All of the sessions are structured in the same way, presenting the following materials:

- _a. A narrated slideshow that demos the content we're working on_
- _b. The slideshow itself, which allows you to click links, copy code, etc._
- _c. A set of training exercises, marked as either 'Homework' or 'Classwork'_
- _d. Solutions to the training exercises, which will come online after class_
<br>
<br>
---

```{r packages, echo=FALSE}
library(vembedr)
```

## a. 0: Demo

```{r, echo = FALSE}
embed_url("https://youtu.be/JORVeMI8HUE")
```
<br>
<br>

## b. 0: Slides

You can access the full slideshow used in the **0-Overview** narration [here](https://simonjbrandl.github.io/mce_xaringan_slides/0-overview.html#1). 
<br>
<br>

## c. 0: Exercises

Normally, this section would have exercises to be done as either **homework** or **classwork**. However, since it's still early in the semester, I'll let you get away without homework on this one! `r emo::ji('winking_face')`
<br>
<br>

## d. 0: Solutions

This section will provide the solutions for homework and classwork exercises. But since there are no exercises, we also don't have any solutions `r emo::ji('pensive_face')`
