---
title: "Session 0: Overview"
description: |
  This session provides a quick overview over the course and a short primer to installing and organizing your R environment. 
author:
  - name: SJB
color1: "#0D2A47"
date: "`r Sys.Date()`"
output:
  postcards::onofre
---

# Let's get started...
<br>
<br>

## Before we start

There are many resources that can help you with learning to code in R. By far the best resource is the vast community of R users and developers and with every issue you run into, you can almost guarantee that someone else had the same question previously. Forums like [stackoverflow](https://stackoverflow.com/) usually have really helpful answers to questions you may have.

One of my favorite formal training modules is the [swirl](https://swirlstats.com/) package, which helps you learn how to code in R by... well, making you code in R. However, before that, you will have to install and organize your R software. The primer below will help you do just that.

## Structure of coding lesson pages

All of the sessions are structured in the same way, presenting the following materials:

- A narrated slideshow that demos the content we're working on
- The slideshow itself, which allows you to click links, copy code, etc.
- A set of training exercises, marked as either 'Homework' or 'Classwork'
- Solutions to the training exercises, which will come online after class

### Narrated slideshow
```{r, echo =FALSE}
library(vembedr)
embed_url("https://youtu.be/JORVeMI8HUE")
```

### Slides

```{r share-again, echo=FALSE}
xaringanExtra::use_share_again()
```

```{r embed-xaringan, echo=FALSE, fig.cap="[Slides](slides/0-overview-x.html)"}
xaringanExtra::embed_xaringan(url = "slides/0-overview-x.html", ratio = "16:9")
```
\
