---
title: "Session 1: Introduction to R"
description: |
  This session provides an introduction to the most basic operations and functions in R. 
author:
  - name: Simon J. Brandl, PhD
    url: https://www.fishandfunctions.com/
    affiliation: The University of Texas at Austin
date: "`r Sys.Date()`"
color1: "#0D2A47"
output:
  distill::distill_article:
    toc: true
    number_sections: true
    toc_depth: 4
    code_folding: false
    toc_float: true
    collapsed: false
    smooth_scroll: true
---

# a. 1: Demo

```{r, echo = FALSE}
library(vembedr)
embed_url("https://youtu.be/kg7mIYvCArU")
```
<br>
<br>

# b. 1: Slides

You can access the full slideshow used in the **1-Introduction** narration **[here](https://simonjbrandl.github.io/mce_xaringan_slides/01-introduction.html#1)**. 

The dataset called 'coralreefherbivores.csv' can be downloaded [here](data/coralreefherbivores.csv).
<br>
<br>

# c. 1: Exercises

### Part I 

- a) create the following vectors:
  - 4 5 6 4 5 6 4 5 6 4 5 6 4 5 6 4 5 6 
  - a sequence from 2 to 36, repeated three times
  - the words "bluefish" "redfish" and "whitefish" in this sequence, six times
  - combine the last two vectors into a new vector
<br>
<br>

- b) create a vector using <span style="color:orange"> *rpois(100, 10)* </span> 
  - determine the frequency of the number 8 in your vector
  - determine how many numbers in your vector are greater than 10
  - determine how many numbers in your vector are smaller than 10 but greater than 5
<br>
<br>

- c) build a matrix that looks like this:

 **[,1]** | **[,2]** | **[,3]** | **[,4]** | **[,5]**
---|---|---|---|---
**[1,]** | 1 | 2 | 3 | 4
**[2,]** | 5 | 6 | 7 | 8
**[3,]** | 9 | 10 | 11 | 12
**[4,]** | 13 | 14 | 15 | 16
**[5,]** | 17 | 18 | 19 | 20 

<br>
<br>

  - create a sequence of multiples of nines between 0 and 100, repeated ten times
<br>
<br>

  - create a vector with random values using <span style="color:orange"> rnorm(1000, 0) </span>, with an explanation of what <span style="color:orange"> rnorm() </span> does
<br>
<br>

  - create a second vector using rnorm, but this time with 10,000 elements, a mean of 50, and a standard deviation of 10
<br>
<br>

  - tell me what the following functions tell you about your vectors: <span style="color:orange"> length(), str(), summary(), head(), tail() </span> 
<br>
<br>

  - retain only the elements from your previous vector that are greater than the mean but smaller than the 3rd Quantile
<br>
<br>

### Part II

- a) create a vector using <span style="color:orange"> *rpois(500, 10)* </span> 
  - determine the proportional frequency of numbers that are >15
  - sort the vector in descending order
  - take the log of the vector and then determine the mean value
  - replace every 10th value in the vector with 0, take the log, and calculate the mean. Exclude values that cause the mean to be -Inf
<br>
<br>

- b) build a matrix that looks like this: 

 **[]** | **redfish** | **bluefish** | **whitefish** | **yellowfish**
---|---|---|---|---
**S1** | 0 | 3 | 6 | 9
**S2** | 12 | 15 | 18 | 21
**S3** | 24 | 27 | 30 | 33
**S4** | 24 | 22 | 20 | 18
**S5** | 16 | 14 | 12 | 10
**S6** | 8 | 6 | 4 | 2
  
  - turn your matrix into a data.frame and add a column entitled "Location" that contains the following sequence: Texas, Belize, Texas, Belize, Texas
  - determine the mean number of fish at site S3
  - determine the difference in the mean number of bluefish and yellowfish across all sites
<br>
<br>

- c) build the following matrix:
  - 100 rows, 20 columns, filled column-wise with randomly-drawn, normally distributed values with a mean of 100 and a standard deviation of 10
  - determine the value of the second-to-last row in column 5
  - determine which column has the largest value in the matrix and which column has the greatest deviation from the mean (100)
<br>
<br>

- d) load in the 'coralreefherbivores.csv' dataset and
  - turn it into a tibble
  - determine the data classes of each column
  - compare the snoutlength of the species _Chlorurus microrhinos_ and _Zebrasoma velifer_
<br>
<br>


# d. 1: Solutions

### Part I

a) 

```{r, echo = FALSE, results = FALSE}
a1<-rep(seq(from=4,to=6,by=1),times=6) %>%
  print()

a2 <- rep(seq(from = 2, to = 36, by = 1), times = 3)
a2

a3 <- rep(c("bluefish", "redfish", "whitefish"), times = 6)
a3

a4 <- paste(a2, a3)
a4
```

b) 

```{r, echo = FALSE, results = FALSE}
b1 <- rpois(100,10)
b1

b2 <- b1[b1 == 8]
table(b2)
length(b2)

b3 <- b1[b1 >10]
b3

b4 <- b1[b1 <10 & b1 >5]
b4
```

c)
```{r, echo = FALSE, results = FALSE}
c1 <- matrix(1:20, nrow = 5, byrow = TRUE)
c1
```
<br>
<br>

### Part II

a)
```{r, echo = FALSE, results = FALSE}
a1 <- rpois(500, 10)
head(a1)

a2 <- a1[a1 > 15]
length(a2)/length(a1)

a3 <- rev(sort(a1))
head(a3)

a4 <- log(a1)
head(a4)

#a5
a1[seq(from=0,to=500,by=10)]<-0
#or
replace(a1, (seq(from=0, to=500, by=10)), 0)

a5logmean <- mean(log(a1[seq(from=0,to=500,by=10)]<-0)) # doesn't work because of -Inf values, so let's go step-by-step

a5step1 <- replace(a1, (seq(from=0, to=500, by=10)), 0) # replace with 0s
a5step2 <- a5step1[a5step1 > 0] # remove the 0s
a5step3 <- mean(log(a5step2)) # take the log of the mean
```

b) 
```{r, echo = FALSE, results = FALSE}
b1 <- matrix(c(seq(from = 0, to = 33, by = 3), seq(from = 24, to = 2, by = -2)), ncol = 4, byrow = TRUE)
rownames(b1) <- c("S1", "S2", "S3", "S4", "S5", "S6")
colnames(b1) <- c("redfish", "bluefish", "whitefish", "yellowfish")

b2 <- as.data.frame(b1) # turn into data frame
b2$Location <- c(rep(c("Texas", "Belize"), times = 3)) # add new column

b3 <- as.matrix(b2[3,c(1:4)]) # turn row 3 and columns 1:4 back into a matrix
mean(b3) # take the mean of the matrix 

print(b4blue <- mean(b2$bluefish)) # mean of bluefish
print(b4yellow <- mean(b2$yellowfish)) # mean of yellowfish
print(b4difference <- b4blue - b4yellow) # difference
```

c) 
```{r, echo = FALSE, results = FALSE}
c1 <- matrix(rnorm(2000, 100, 10), ncol = 20, byrow = F)
head(c1)

c2 <- c1[98, 5] # look for value in row 98 column 5
c2

summary(c1) # look for maximum value and largest deviation from the specified mean (100)
```

d) 
```{r, echo = FALSE, results = FALSE}
herbivores <- read.csv(file = "data/coralreefherbivores.csv")
herbivore.tbl <- dplyr::as_tibble(herbivores) # turn into tibble - might need a new package here ;-)
herbivore.tbl # tibbles tell you the data classes but str() would also work well
herbivore.tbl$gen.spe # the two species are rows 31 and 96
herbivore.tbl[c(31,96),] # check snoutlength
```