---
title: "Session 3: Data visualization in ggplot2"
description: |
  This session introduces the basic functionalities of the ggplot2 package, allowing you to create beautiful graphs using R. 
author:
  - name: Simon J. Brandl, PhD
    url: https://www.fishandfunctions.com/
    affiliation: The University of Texas at Austin
date: "`r Sys.Date()`"
color1: "#0D2A47"
output:
  distill::distill_article:
    toc: true
    number_sections: true
    toc_depth: 4
    code_folding: false
    toc_float: true
    collapsed: false
    smooth_scroll: true
editor_options: 
  chunk_output_type: console
---

```{r, echo = F}
library(tidyverse)
library(fishualize)
library(ggridges)
```
# a. 3: Demo

```{r, echo = FALSE}
library(vembedr)
embed_url("https://youtu.be/gyNf7-KdBDI")
```
<br>
<br>

# b. 3: Slides

You can access the full slideshow used in the **3-ggplot** narration **[here](https://simonjbrandl.github.io/mce_xaringan_slides/03-ggplot.html)**.

The dataset called 'coralreefherbivores.csv' can be downloaded [here](data/coralreefherbivores.csv).

The dataset called 'reef_fishes.csv' can be downloaded [here](data/reef_fishes.csv).

The dataset called 'fish_abundance.csv' can be downloaded [here](data/fish_abundance.csv).

The dataset called 'site_info.csv' can be downloaded [here](data/site_info.csv).
<br>
<br>


# c. 3: Exercises

## Homework

_Please complete the following exercises at home after going through the demo above. We will go through the solutions to these exercises in class on Wednesday Feb 7th_

Start by loading in the two datasets below (which are available for download using the links above).

```{r, echo = TRUE}
fish.abundance <- read.csv(file = "data/fish_abundance.csv")
site.info <- read.csv(file = "data/site_info.csv")
```

Re-familiarize with the "fish_abundance.csv" dataset. The dataset includes many fish species, which were counted across different sites and depths. Fish diversity often decreases with depth on coral reefs, so let's explore whether there is a relationship between depth and diversity.

### Part I

1) Simplify the dataset to include species, site, depth.
2) Use distinct() to make sure no species were recorded twice.
3) Calculate species richness by site and depth.

### Part II

1) Plot depth versus species richness.
2) Add site as a color variable.

### Part III

1) Produce another plot, only including the families Serranidae, Acanthuridae, Pomacentridae, and Chaetodontidae.
2) Use with facet_wrap() or facet_grid() to create separate facets for each family.

### Bonus work `r emo::ji('star_struck')`

1) Explore differences in the species richness of the four families (Serranidae, Acanthuridae, Pomacentridae, and Chaetodontidae) using violin or density plots
2) Color the density plots based on family and include the raw data to create sina plots
3) Try to sort the violin plot in descending order, starting with the highest species richness
4) Make it pretty! `r emo::ji('blush')`

## Classwork

_We will do these exercises in class on Wednesday, February 7th_

In the last plot from Exercise 1, it appears as though some sites have higher species richness than others. Let's further examine why species richness across sites using "site_info.csv," which includes metadata on the exposure of each site. 

### Part I

1) Create a dataset that summarizes species richness by surveyid and site.
2) Join this species richness data with the site exposure metadata.

### Part II

1) Use a violin plot to visualize the species richness of exposed vs. lagoon sites.

### Part III

1) Calculate the average abundance of each family in a given survey.
2) Plot the average abundance using density curves.
3) Bonus: Transform the x-axis to make the plot more useful.

# d. 3: Solutions

## Homework solutions

### Part I

1) Simplify the dataset to include species, site, depth.
2) Use distinct() to make sure no species were recorded twice.
3) Calculate species richness by site and depth.

```{r, echo = TRUE}

fish.abundance <- read.csv(file = "data/fish_abundance.csv") # load data

fish.sprich <- fish.abundance %>%
  select(site, depth, genspe) %>%
  distinct() %>%
  group_by(site, depth) %>%
  summarize(sprich = n())

head(fish.sprich)
```

### Part II

1) Plot depth versus species richness.
2) Add site as a color variable.

```{r, echo = TRUE}

fish.sprich.plot <- ggplot(fish.sprich, aes(x = depth, y = sprich, color = site)) +
  geom_point() +
  scale_color_fish_d(option = "Synchiropus_splendidus") +
  theme_bw()
  
fish.sprich.plot
```

### Part III

1) Produce another plot, only including the families Serranidae, Acanthuridae, Pomacentridae, and Chaetodontidae.
2) Use with facet_wrap() or facet_grid() to create separate facets for each family.

```{r, echo = TRUE}

fish.sprich.fam <- fish.abundance %>%
  filter(family %in% c("Acanthuridae", "Chaetodontidae", "Serranidae", "Pomacentridae")) %>%
  select(site, depth, family, genspe) %>%
  distinct() %>%
  group_by(site, depth, family) %>%
  summarize(sprich = n()) 
head(fish.sprich.fam)

fish.sprich.fam.plot <- ggplot(fish.sprich.fam, aes(x = depth, y = sprich, color = site)) +
  geom_point() + 
  facet_wrap(.~family, scales = "free") +
    scale_color_fish_d(option = "Synchiropus_splendidus") +
  theme_bw()
fish.sprich.fam.plot
```

## Bonuswork solutions

1) Explore differences in the species richness of the four families (Serranidae, Acanthuridae, Pomacentridae, and Chaetodontidae) using violin or density plots
2) Color the density plots based on family and include the raw data to create sina plots
3) Try to sort the violin plot in descending order, starting with the highest species richness
4) Make it pretty! `r emo::ji('blush')`

```{r, echo = TRUE}
fish.family.plot <- ggplot(fish.sprich.fam, aes(x = fct_reorder(family, -sprich, .fun = sum), 
                                                y = sprich, fill = family)) +
  geom_violin(draw_quantiles = c(0.05, 0.5, 0.95), color = "grey23", alpha = 0.8, lwd = 0.5) +
  geom_jitter(aes(shape = family), alpha = 0.5, color = "black", size = 3, width = 0.2) +
  theme_bw() +
  scale_fill_fish_d(option = "Bodianus_rufus", name = "Fish family") +
  scale_shape_manual(values = c(21:24), name = "Fish family") +
  ylab("Number of species per site") +
  xlab("Fish family") 
  
fish.family.plot
```


## Classwork solutions

## Part I

1) Create a dataset that summarizes species richness by surveyid and site.
2) Join this species richness data with the site exposure metadata.

```{r, echo = TRUE}
site.info <- read.csv(file = "data/site_info.csv")

fish.sprich.site <- fish.abundance %>%
  select(surveyid, site, genspe) %>%
  distinct() %>%
  group_by(surveyid, site) %>%
  summarize(sprich = n()) %>%
  left_join(site.info, by = "site")

head(fish.sprich.site) 
```

## Part II

1) Use a violin plot to visualize the species richness of exposed vs. lagoon sites.

```{r, echo = TRUE}
fish.sprich.site.plot <- ggplot(fish.sprich.site, aes(x = exposure, y = sprich, fill = exposure)) +
  geom_violin(draw_quantiles = c(0.025, 0.5, 0.975)) +
  geom_jitter(width = 0.1) +
    scale_fill_fish_d(option = "Synchiropus_splendidus") +
  theme_bw()
fish.sprich.site.plot
```

## Part III

1) Calculate the average abundance of each family in a given survey.
2) Plot the average abundance using density curves.
3) Bonus: transform the x-axis to make the plot more useful.

```{r, echo = TRUE}
# 1) Calculate the average abundance of each family in a given survey.
fish.abun.survey <- fish.abundance %>%
  group_by(surveyid, family) %>%
  summarize(total.fish = sum(total))
head(fish.abun.survey)

# 2) Plot the average abundance using density curves.
fish.abun.plot <- ggplot(fish.abun.survey, 
                        aes(x = total.fish, y = family)) +
                          geom_density_ridges(alpha = 0.5, fill = "forestgreen") +
  theme_bw()
fish.abun.plot

# 3) Bonus: Transform the x-axis to make the plot more useful.
# use rel_min_height() to cut the tails
fish.abun.plot2 <- ggplot(fish.abun.survey, 
                        aes(x = log10(total.fish), #use log10 transformation
                            y = fct_reorder(family, total.fish, .fun = sum))) + # use fct_reorder() to reorder the y-variable as descending based on the total sum of fish in each family
                          geom_density_ridges(alpha = 0.5, rel_min_height = 0.005, fill = "forestgreen") +
  xlab("Species richness (log)") +
  ylab("Fish family") +
  theme_bw()
fish.abun.plot2
```
