---
title: "Session 4: Modeling ecological processes"
description: |
  This session introduces for loops as means to perform numerical simulations in R and allows you to play with models to simulate ecological processes. 
author:
  - name: Simon J. Brandl, PhD
    url: https://www.fishandfunctions.com/
    affiliation: The University of Texas at Austin
date: "`r Sys.Date()`"
color1: "#0D2A47"
output:
  distill::distill_article:
    toc: true
    number_sections: true
    toc_depth: 4
    code_folding: false
    toc_float: true
    collapsed: false
    smooth_scroll: true
editor_options: 
  chunk_output_type: console
---


```{r, echo = F}
library(tidyverse)
library(fishualize)
library(ggridges)
```
# a. 4: Demo

```{r, echo = FALSE}
library(vembedr)
embed_url("https://youtu.be/bHTRZZBh6yc")
```
<br>
<br>

# b. 4: Slides

You can access the full slideshow used in the **4-Simulations** narration **[here](https://simonjbrandl.github.io/mce_xaringan_slides/04-simulations.html)**.
<br>
<br>

# c. 3: Exercises

## Homework

_If you want to, complete the following exercises at home after going through the demo above. We will go through the solutions to these exercises in class on Wednesday Feb 28th_


- Start by loading the tidyverse package and copy-and-pasting the basic Moran Model into your R. The code for the model is below: 

```{r, echo = TRUE}
library(tidyverse)

num.sims <- 20 # specify the number of simulations 
num.years <- 50 # specify the number of years
freq.1.mat <- matrix(ncol = num.sims, nrow = num.years) # create a matrix for output

# FIRST LOOP FOR DIFFERENT SIMULATIONS
for (j in 1:num.sims) # use for-loop to run through the number of simulations
  { 
  
  J <- 100 # 100 individuals
  t0.sp1 <- 0.5*J  # abundance of species 1 at time 0 
  community <- vector(length = J) # set up community
  community[1:t0.sp1] <- 1 # specify that half of the community is species 1
  community[(t0.sp1+1):J] <- 2 # specify the other half is species 2
  year <- 2 # set 'year' to 2 
  freq.1.mat[1,j] <- sum(community==1)/J # fill community matrix
  
  # SECOND LOOP FOR INDIVIDUAL DEATH/BIRTH
  for (i in 1:(J*(num.years-1))) # second for-loop to run multiple simulations
    { 
    
    freq.1 <- sum(community==1)/J # freq.1 represents the frequency of species 1
    pr.1 <- freq.1 # pr.1 represents the frequency of reproduction
    community[ceiling(J*runif(1))] <- sample(c(1,2), 1, prob=c(pr.1,1-pr.1))  # birth and death rates, based on probabilities
    if (i %% J == 0)  # record data in the freq.1.mat matrix
      { 
      
      freq.1.mat[year, j] <- sum(community==1)/J
      year <- year + 1
      
    }
  }
}
```


- After that, wrangle the data as done in the lecture and produce the beautiful graph below. 

```{r, echo = TRUE}
colnames(freq.1.mat) <- 1:num.sims # set column names in matrix

freq.sp1.df <- as.data.frame(freq.1.mat) %>% # convert freq.1.mat into data frame 
  add_column(year = 1:num.years) %>% # add a column called year
  gather(1:20, key = iteration, value = frequency) # gather into long format

p1 <- ggplot(freq.sp1.df, 
             aes(x = year, 
                 y = frequency, 
                 group = iteration)) +
  geom_line(aes(color = frequency), 
            alpha = 0.8) +
  scale_color_gradient(low = "darkorchid",
                       high = "goldenrod") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(0,1)) +
  xlab("Years") +
  ylab("Frequency of species 1")

p1
```

- Accomplished? Well done. Here are your exercises:

### Part I:

1) Run the same model but change the _number of individuals to 500_ and _increase the number of years to 75_

2) Run the same model but decrease the _number of individuals to 50_, set the _number of years back to 50_, and _increase the number of simulations to 50_. Be careful with changing the number of simulations: you will have to adjust your downstream data wrangling code in the _gather()_ function. 

3) Change the colors of your lines to be a beautiful green and blue instead and increase the opacity to 0.9.


## Classwork

### Part I

Introduce a selective advantage for Species 1 using a fitness ratio of 1.2. The fitness ratio affects the probability of reproduction in the following terms: fit.ratio * freq.1 / (fit.ratio * freq.1 + freq.2) 

1) Run a model with 300 individuals, 30 simulations, 50 years, and a fitness ratio of 1.2. Plot your outcomes (and make it pretty).

2) Run the same model with 1000 individuals, 30 simulations, 50 years, and a fitness ratio of 1.2. Plot your outcomes next to your previous plot.

### Part II

Introduce negative frequency dependence into your model. To do so, you first need to define your average fitness ratio (fit.ratio.avg) based on the prevailing selective advantage. Frequency dependence then affects the fit.ratio in the following terms: exp(freq.dep * (freq.1 - 0.5) + log(fit.ratio.avg)). 

1) Run a model with 300 individuals, 30 simulations, 50 years, an average fitness ratio of 1.2, and a negative frequency dependence of -0.3. Plot your outcomes. 

2) Run a model with 1000 individuals, 30 simulations, 50 years, a fitness ratio of 1.5, and a negative frequency dependence of -0.5. Plot your outcomes. 

### Part III

Introduce positive frequency dependence and change your starting densities. To do so, follow the same steps as above but also adjust your initial populations.

1) Run a model with 1000 individuals, 30 simulations, 50 years, a fitness ratio of 1.1, and a positive frequency dependence of 0.1, with a starting frequency of Species 1 at 0.3.

2) Run a model with 1000 individuals, 30 simulations, 50 years, a fitness ratio of 1.0, a positive frequency dependence of 0.3, and a starting frequency of Species 1 at 0.75.

# d. 4: Solutions

## Homework solutions

### Part I:

1) Run the same model but change the _number of individuals to 500_ and _increase the number of years to 75_

```{r, echo = TRUE}
library(tidyverse)

num.sims <- 30 # specify the number of simulations 
num.years <- 75 # specify the number of years
freq.1.mat <- matrix(ncol = num.sims, nrow = num.years) # create a matrix for output

# FIRST LOOP FOR DIFFERENT SIMULATIONS
for (j in 1:num.sims) # use for-loop to run through the number of simulations
  { 
  
  J <- 500 # 500 individuals
  t0.sp1 <- 0.5*J  # abundance of species 1 at time 0 
  community <- vector(length = J) # set up community
  community[1:t0.sp1] <- 1 # specify that half of the community is species 1
  community[(t0.sp1+1):J] <- 2 # specify the other half is species 2
  year <- 2 # set 'year' to 2 
  freq.1.mat[1,j] <- sum(community==1)/J # fill community matrix
  
  # SECOND LOOP FOR INDIVIDUAL DEATH/BIRTH
  for (i in 1:(J*(num.years-1))) # second for-loop to run multiple simulations
    { 
    
    freq.1 <- sum(community==1)/J # freq.1 represents the frequency of species 1
    pr.1 <- freq.1 # pr.1 represents the frequency of reproduction
    community[ceiling(J*runif(1))] <- sample(c(1,2), 1, prob=c(pr.1,1-pr.1))  # birth and death rates, based on probabilities
    if (i %% J == 0)  # record data in the freq.1.mat matrix
      { 
      
      freq.1.mat[year, j] <- sum(community==1)/J
      year <- year + 1
      
    }
  }
}

colnames(freq.1.mat) <- 1:num.sims # set column names in matrix

freq.sp1.df <- as.data.frame(freq.1.mat) %>% # convert freq.1.mat into data frame 
  add_column(year = 1:num.years) %>% # add a column called year
  gather(1:30, key = iteration, value = frequency) # gather into long format

p1 <- ggplot(freq.sp1.df, 
             aes(x = year, 
                 y = frequency, 
                 group = iteration)) +
  geom_line(aes(color = frequency), 
            alpha = 0.8) +
  scale_color_gradient(low = "darkorchid",
                       high = "goldenrod") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(0,1)) +
  xlab("Years") +
  ylab("Frequency of species 1")
p1
```

2) Run the same model but decrease the _number of individuals to 50_, set the _number of years back to 50_, and _increase the number of simulations to 50_. Be careful with changing the number of simulations: you will have to adjust your downstream data wrangling code in the _gather()_ function. 

```{r, echo = TRUE}
library(tidyverse)

num.sims <- 50 # specify the number of simulations 
num.years <- 50 # specify the number of years
freq.1.mat <- matrix(ncol = num.sims, nrow = num.years) # create a matrix for output

# FIRST LOOP FOR DIFFERENT SIMULATIONS
for (j in 1:num.sims) # use for-loop to run through the number of simulations
  { 
  
  J <- 50 # 500 individuals
  t0.sp1 <- 0.5*J  # abundance of species 1 at time 0 
  community <- vector(length = J) # set up community
  community[1:t0.sp1] <- 1 # specify that half of the community is species 1
  community[(t0.sp1+1):J] <- 2 # specify the other half is species 2
  year <- 2 # set 'year' to 2 
  freq.1.mat[1,j] <- sum(community==1)/J # fill community matrix
  
  # SECOND LOOP FOR INDIVIDUAL DEATH/BIRTH
  for (i in 1:(J*(num.years-1))) # second for-loop to run multiple simulations
    { 
    
    freq.1 <- sum(community==1)/J # freq.1 represents the frequency of species 1
    pr.1 <- freq.1 # pr.1 represents the frequency of reproduction
    community[ceiling(J*runif(1))] <- sample(c(1,2), 1, prob=c(pr.1,1-pr.1))  # birth and death rates, based on probabilities
    if (i %% J == 0)  # record data in the freq.1.mat matrix
      { 
      
      freq.1.mat[year, j] <- sum(community==1)/J
      year <- year + 1
      
    }
  }
}

colnames(freq.1.mat) <- 1:num.sims # set column names in matrix

freq.sp1.df <- as.data.frame(freq.1.mat) %>% # convert freq.1.mat into data frame 
  add_column(year = 1:num.years) %>% # add a column called year
  gather(1:30, key = iteration, value = frequency) # gather into long format

p1 <- ggplot(freq.sp1.df, 
             aes(x = year, 
                 y = frequency, 
                 group = iteration)) +
  geom_line(aes(color = frequency), 
            alpha = 0.8) +
  scale_color_gradient(low = "darkorchid",
                       high = "goldenrod") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(0,1)) +
  xlab("Years") +
  ylab("Frequency of species 1")
p1
```

3) Change the colors of your lines to be a beautiful green and blue instead and increase the opacity to 0.9.

```{r, echo = TRUE}
p1 <- ggplot(freq.sp1.df, 
             aes(x = year, 
                 y = frequency, 
                 group = iteration)) +
  geom_line(aes(color = frequency), 
            alpha = 0.9) +
  scale_color_gradient(low = "forestgreen",
                       high = "steelblue") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(0,1)) +
  xlab("Years") +
  ylab("Frequency of species 1")
p1
```

## Classwork solutions

### Part I

Introduce a selective advantage for Species 1 using a fitness ratio of 1.2. The fitness ratio affects the probability of reproduction in the following terms: fit.ratio * freq.1 / (fit.ratio * freq.1 + freq.2) 

1) Run a model with 300 individuals, 30 simulations, 50 years, and a fitness ratio of 1.2. Plot your outcomes (and make it pretty).

```{r, echo = TRUE}
# selection
num.sims <- 30 
num.years <- 50
freq.1.mat <- matrix(ncol = num.sims, nrow = num.years)

for (j in 1:num.sims) {
  J <- 300
  t0.sp1 <- 0.5*J  
  community <- vector(length = J)
  community[1:t0.sp1] <- 1
  community[(t0.sp1+1):J] <- 2
  year <- 2
  freq.1.mat[1,j] <- sum(community==1)/J
  
  for (i in 1:(J*(num.years-1))) {
    freq.1 <- sum(community==1)/J 
    freq.2 <- 1 - freq.1
    fit.ratio <- 1.2 
    pr.1 <- fit.ratio*freq.1/(fit.ratio*freq.1 + freq.2) 
    community[ceiling(J*runif(1))] <- sample(c(1,2), 1, prob=c(pr.1,1-pr.1)) 
    
    # record data in the freq.1.mat matrix   
    if (i %% J == 0) {
      freq.1.mat[year, j] <- sum(community==1)/J
      year <- year + 1
    }
  }
}

colnames(freq.1.mat) <- 1:num.sims # set column names in matrix

freq.sp1.df <- as.data.frame(freq.1.mat) %>% # convert freq.1.mat into data frame 
  add_column(year = 1:num.years) %>% # add a column called year
  gather(1:30, key = iteration, value = frequency) # gather into long format

p1 <- ggplot(freq.sp1.df, 
             aes(x = year, 
                 y = frequency, 
                 group = iteration)) +
  geom_line(aes(color = frequency), 
            alpha = 0.8) +
  scale_color_gradient(low = "firebrick",
                       high = "turquoise") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(0,1)) +
  xlab("Years") +
  ylab("Frequency of species 1")
p1
```

2) Run the same model with 1000 individuals, 30 simulations, 50 years, and a fitness ratio of 1.2. Plot your outcomes next to your previous plot.

```{r, echo = TRUE}
# selection
num.sims <- 30 
num.years <- 50
freq.1.mat <- matrix(ncol = num.sims, nrow = num.years)

for (j in 1:num.sims) {
  J <- 1000
  t0.sp1 <- 0.5*J  
  community <- vector(length = J)
  community[1:t0.sp1] <- 1
  community[(t0.sp1+1):J] <- 2
  year <- 2
  freq.1.mat[1,j] <- sum(community==1)/J
  
  for (i in 1:(J*(num.years-1))) {
    freq.1 <- sum(community==1)/J 
    freq.2 <- 1 - freq.1
    fit.ratio <- 1.2 
    pr.1 <- fit.ratio*freq.1/(fit.ratio*freq.1 + freq.2) 
    community[ceiling(J*runif(1))] <- sample(c(1,2), 1, prob=c(pr.1,1-pr.1)) 
    
    # record data in the freq.1.mat matrix   
    if (i %% J == 0) {
      freq.1.mat[year, j] <- sum(community==1)/J
      year <- year + 1
    }
  }
}

colnames(freq.1.mat) <- 1:num.sims # set column names in matrix

freq.sp1.df <- as.data.frame(freq.1.mat) %>% # convert freq.1.mat into data frame 
  add_column(year = 1:num.years) %>% # add a column called year
  gather(1:30, key = iteration, value = frequency) # gather into long format

p1 <- ggplot(freq.sp1.df, 
             aes(x = year, 
                 y = frequency, 
                 group = iteration)) +
  geom_line(aes(color = frequency), 
            alpha = 0.8) +
  scale_color_gradient(low = "firebrick",
                       high = "turquoise") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(0,1)) +
  xlab("Years") +
  ylab("Frequency of species 1")
p1
```

### Part II

Introduce negative frequency dependence into your model. To do so, you first need to define your average fitness ratio (fit.ratio.avg) based on the prevailing selective advantage. Frequency dependence then affects the fit.ratio in the following terms: exp(freq.dep * (freq.1 - 0.5) + log(fit.ratio.avg)). 

1) Run a model with 300 individuals, 30 simulations, 50 years, an average fitness ratio of 1.2, and a negative frequency dependence of -0.3. Plot your outcomes. 

```{r, echo = FALSE}
# negative freq dep
num.sims <- 30 
num.years <- 50
freq.1.mat <- matrix(ncol = num.sims, nrow = num.years)

freq.dep = -0.3
fit.ratio.avg <- 1.2 

for (j in 1:num.sims) {
  J <- 300
  t0.sp1 <- 0.5*J  
  community <- vector(length = J)
  community[1:t0.sp1] <- 1
  community[(t0.sp1+1):J] <- 2
  year <- 2
  freq.1.mat[1,j] <- sum(community==1)/J
  
  for (i in 1:(J*(num.years-1))) {
    freq.1 <- sum(community==1)/J 
    freq.2 <- 1 - freq.1
    fit.ratio <- exp(freq.dep*(freq.1-0.5) + log(fit.ratio.avg))
    pr.1 <- fit.ratio*freq.1/(fit.ratio*freq.1 + freq.2)
    community[ceiling(J*runif(1))] <- sample(c(1,2), 1, prob=c(pr.1,1-pr.1)) 
    
    # record data in the freq.1.mat matrix   
    if (i %% J == 0) {
      freq.1.mat[year, j] <- sum(community==1)/J
      year <- year + 1
    }
  }
}

colnames(freq.1.mat) <- 1:num.sims # set column names in matrix

freq.sp1.df <- as.data.frame(freq.1.mat) %>% # convert freq.1.mat into data frame 
  add_column(year = 1:num.years) %>% # add a column called year
  gather(1:30, key = iteration, value = frequency) # gather into long format

p1 <- ggplot(freq.sp1.df, 
             aes(x = year, 
                 y = frequency, 
                 group = iteration)) +
  geom_line(aes(color = frequency), 
            alpha = 0.8) +
  scale_color_gradient(low = "cyan3",
                       high = "bisque3") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(0,1)) +
  xlab("Years") +
  ylab("Frequency of species 1")
p1
```

2) Run a model with 1000 individuals, 30 simulations, 50 years, a fitness ratio of 1.5, and a negative frequency dependence of -0.5. Plot your outcomes. 

```{r, echo = FALSE}
# negative freq dep
num.sims <- 30 
num.years <- 50
freq.1.mat <- matrix(ncol = num.sims, nrow = num.years)

freq.dep = -0.5
fit.ratio.avg <- 1.5

for (j in 1:num.sims) {
  J <- 1000
  t0.sp1 <- 0.5*J  
  community <- vector(length = J)
  community[1:t0.sp1] <- 1
  community[(t0.sp1+1):J] <- 2
  year <- 2
  freq.1.mat[1,j] <- sum(community==1)/J
  
  for (i in 1:(J*(num.years-1))) {
    freq.1 <- sum(community==1)/J 
    freq.2 <- 1 - freq.1
    fit.ratio <- exp(freq.dep*(freq.1-0.5) + log(fit.ratio.avg))
    pr.1 <- fit.ratio*freq.1/(fit.ratio*freq.1 + freq.2)
    community[ceiling(J*runif(1))] <- sample(c(1,2), 1, prob=c(pr.1,1-pr.1)) 
    
    # record data in the freq.1.mat matrix   
    if (i %% J == 0) {
      freq.1.mat[year, j] <- sum(community==1)/J
      year <- year + 1
    }
  }
}

colnames(freq.1.mat) <- 1:num.sims # set column names in matrix

freq.sp1.df <- as.data.frame(freq.1.mat) %>% # convert freq.1.mat into data frame 
  add_column(year = 1:num.years) %>% # add a column called year
  gather(1:30, key = iteration, value = frequency) # gather into long format

p1 <- ggplot(freq.sp1.df, 
             aes(x = year, 
                 y = frequency, 
                 group = iteration)) +
  geom_line(aes(color = frequency), 
            alpha = 0.8) +
  scale_color_gradient(low = "cyan3",
                       high = "bisque3") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(0,1)) +
  xlab("Years") +
  ylab("Frequency of species 1")
p1
```

### Part III

Introduce positive frequency dependence and change your starting densities. To do so, follow the same steps as above but also adjust your initial populations.

1) Run a model with 1000 individuals, 30 simulations, 50 years, a fitness ratio of 1.1, and a positive frequency dependence of 0.1, with a starting frequency of Species 1 at 0.3.

```{r, echo = TRUE}
# positive frequency dependence
num.sims <- 30 
num.years <- 50
freq.1.mat <- matrix(ncol = num.sims, nrow = num.years)

freq.dep = 0.1
fit.ratio.avg <- 1.1

for (j in 1:num.sims) {
  J <- 1000
  t0.sp1 <- 0.3*J  
  community <- vector(length = J)
  community[1:t0.sp1] <- 1
  community[(t0.sp1+1):J] <- 2
  year <- 2
  freq.1.mat[1,j] <- sum(community==1)/J
  
  for (i in 1:(J*(num.years-1))) {
    freq.1 <- sum(community==1)/J 
    freq.2 <- 1 - freq.1
    fit.ratio <- exp(freq.dep*(freq.1-0.5) + log(fit.ratio.avg))
    pr.1 <- fit.ratio*freq.1/(fit.ratio*freq.1 + freq.2)
    community[ceiling(J*runif(1))] <- sample(c(1,2), 1, prob=c(pr.1,1-pr.1)) 
    
    # record data in the freq.1.mat matrix   
    if (i %% J == 0) {
      freq.1.mat[year, j] <- sum(community==1)/J
      year <- year + 1
    }
  }
}

colnames(freq.1.mat) <- 1:num.sims # set column names in matrix

freq.sp1.df <- as.data.frame(freq.1.mat) %>% # convert freq.1.mat into data frame 
  add_column(year = 1:num.years) %>% # add a column called year
  gather(1:30, key = iteration, value = frequency) # gather into long format

p1 <- ggplot(freq.sp1.df, 
             aes(x = year, 
                 y = frequency, 
                 group = iteration)) +
  geom_line(aes(color = frequency), 
            alpha = 0.8) +
  scale_color_gradient(low = "darkviolet",
                       high = "darkseagreen1") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(0,1)) +
  xlab("Years") +
  ylab("Frequency of species 1")
p1
```

2) Run a model with 1000 individuals, 30 simulations, 50 years, a fitness ratio of 1.0, a positive frequency dependence of 0.3, and a starting frequency of Species 1 at 0.75.

```{r, echo = TRUE}
# positive frequency dependence
num.sims <- 30 
num.years <- 50
freq.1.mat <- matrix(ncol = num.sims, nrow = num.years)

freq.dep = 0.3
fit.ratio.avg <- 1.0

for (j in 1:num.sims) {
  J <- 1000
  t0.sp1 <- 0.75*J  
  community <- vector(length = J)
  community[1:t0.sp1] <- 1
  community[(t0.sp1+1):J] <- 2
  year <- 2
  freq.1.mat[1,j] <- sum(community==1)/J
  
  for (i in 1:(J*(num.years-1))) {
    freq.1 <- sum(community==1)/J 
    freq.2 <- 1 - freq.1
    fit.ratio <- exp(freq.dep*(freq.1-0.5) + log(fit.ratio.avg))
    pr.1 <- fit.ratio*freq.1/(fit.ratio*freq.1 + freq.2)
    community[ceiling(J*runif(1))] <- sample(c(1,2), 1, prob=c(pr.1,1-pr.1)) 
    
    # record data in the freq.1.mat matrix   
    if (i %% J == 0) {
      freq.1.mat[year, j] <- sum(community==1)/J
      year <- year + 1
    }
  }
}

colnames(freq.1.mat) <- 1:num.sims # set column names in matrix

freq.sp1.df <- as.data.frame(freq.1.mat) %>% # convert freq.1.mat into data frame 
  add_column(year = 1:num.years) %>% # add a column called year
  gather(1:30, key = iteration, value = frequency) # gather into long format

p1 <- ggplot(freq.sp1.df, 
             aes(x = year, 
                 y = frequency, 
                 group = iteration)) +
  geom_line(aes(color = frequency), 
            alpha = 0.8) +
  scale_color_gradient(low = "darkviolet",
                       high = "darkseagreen1") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(0,1)) +
  xlab("Years") +
  ylab("Frequency of species 1")
p1
```