---
title: "Session 5: Principal Component Analysis"
description: |
  This session introduces Principal Component Analyses, their interpretation, visualization, and other ordination types.  
author:
  - name: Simon J. Brandl, PhD
    url: https://www.fishandfunctions.com/
    affiliation: The University of Texas at Austin
date: "`r Sys.Date()`"
color1: "#0D2A47"
output:
  distill::distill_article:
    toc: true
    number_sections: true
    toc_depth: 4
    code_folding: false
    toc_float: true
    collapsed: false
    smooth_scroll: true
editor_options: 
  chunk_output_type: console
---


```{r, echo = F}
library(tidyverse)
library(fishualize)
library(ggridges)
```
# a. 5: Demo

```{r, echo = FALSE}
library(vembedr)
embed_url("https://youtu.be/RcpJ6stmTmA")
```
<br>
<br>

# b. 5: Slides

You can access the full slideshow used in the **5-PCA** narration **[here](https://simonjbrandl.github.io/mce_xaringan_slides/05-PCA.html)**.

The dataset called 'coralreefherbivores.csv' can be downloaded [here](data/coralreefherbivores.csv).

The dataset called 'fishcoms_lizardisland.csv' can be downloaded [here](data/fishcoms_lizardisland.csv).
<br>
<br>

# c. 5: Exercises

- Start by installing and loading the *vegan* and the *ape* package into your environment, along with *dplyr*, *ggplot* and *patchwork* (or load the former two via the *tidyverse* package)

- Then, read in the fishcoms_lizardisland dataset, which is available above

### Part I

- Retain only species in the family Chaetodontidae (butterflyfishes) in the dataset, and summarize each species' abundance across sites, while also retaining information on each site's exposure level. Bring the data to wide format.

```{r, echo = FALSE, results = FALSE}
library(vegan)
library(tidyverse)
library(ape)
library(patchwork)
# re-familiarize with the Lizard Island fish community dataset
lfish <- read.csv(file =  "data/fishcoms_lizardisland.csv")
head(lfish)
lizard.sum <- lfish %>%
  filter(Family == "Chaetodontidae") %>%
  group_by(Site, Exposure, species) %>%
  summarize(abundance = sum(abundance))
lizard.sum

# spread into wide format and get rid of NAs
lizard.wide <- lizard.sum %>%
  spread(key = species, value = abundance, fill = 0)
lizard.wide
```

- Divide your dataset into metadata and abundance data, run a PCA on the Hellinger transformed data, and plot the data using ggplot. See how the data changes depending on whether you scale and center your data. 

```{r, echo = FALSE, results = FALSE}
# isolate metadata
lizard.meta <- lizard.wide[1:2]
lizard.meta

# isolate numeric data
lizard.raw <- lizard.wide %>%
  ungroup() %>%
  select(-c(1:2))
lizard.raw

# the decostand() function in vegan allows for transformations like Hellinger
lizard.hell <- decostand(lizard.raw, method = "hellinger")
head(lizard.hell)

# run a PCA and retain percent variation explained by PC1 and PC2
lizard.pca <- summary(rda(lizard.hell, scale = T))
lizard.pca

# extract the site scores and merge with metadata
lizard.scores <- as.data.frame(lizard.pca$sites) %>%
  bind_cols(lizard.meta)

# create convex hulls grouped by site exposure based on site scores
lizard.hulls <- lizard.scores %>%
  group_by(Exposure) %>%
  slice(chull(PC1, PC2))

# extract the loadings from species scores
lizard.loadings <- as.data.frame(lizard.pca$species) %>%
  mutate(species = rownames(.))

# plot the PCA results with color and convex hulls as exposure and species as loadings
lizard.plot2 <- ggplot() +
  geom_point(data = lizard.scores, aes(x = PC1, y = PC2, color = Exposure), size = 2) +
  geom_polygon(data = lizard.hulls, aes(x = PC1, PC2, fill = Exposure), alpha = 0.5) +
  geom_segment(data = lizard.loadings, aes(x = 0, xend = PC1,
                                           y = 0, yend = PC2), lwd = 0.1) +
  geom_label(data = lizard.loadings, aes(x = PC1, y = PC2, label = species), size = 3) +
  scale_fill_viridis_d() +
  scale_color_viridis_d() +
  theme_bw() +
  xlab("PC1 (19.5%)") +
  ylab("PC2 (14.4%)") +
  ggtitle("Hellinger transformed PCA, scale = T")
lizard.plot2

# run the same PCA, but with scale = F
# retain percent variation explained by PC1 and PC2
lizard.pca <- summary(rda(lizard.hell, scale = F))

# extract the site scores and merge with metadata
lizard.scores <- as.data.frame(lizard.pca$sites) %>%
  bind_cols(lizard.meta)

# create convex hulls grouped by site exposure based on site scores
lizard.hulls <- lizard.scores %>%
  group_by(Exposure) %>%
  slice(chull(PC1, PC2))

# extract the loadings from species scores
lizard.loadings <- as.data.frame(lizard.pca$species) %>%
  mutate(species = rownames(.))

# plot the PCA results with color and convex hulls as exposure and species as loadings
lizard.plot3 <- ggplot() +
  geom_point(data = lizard.scores, aes(x = PC1, y = PC2, color = Exposure), size = 2) +
  geom_polygon(data = lizard.hulls, aes(x = PC1, PC2, fill = Exposure), alpha = 0.5) +
  geom_segment(data = lizard.loadings, aes(x = 0, xend = PC1,
                                           y = 0, yend = PC2), lwd = 0.1) +
  geom_label(data = lizard.loadings, aes(x = PC1, y = PC2, label = species), size = 3) +
  scale_fill_viridis_d() +
  scale_color_viridis_d() +
  theme_bw() +
  xlab("PC1 (22.8%)") +
  ylab("PC2 (18.2%)") +
  ggtitle("Hellinger transformed PCA, scale = F")
lizard.plot3
```

Part II 

- We now want to examine how the four regimes of wave action impact reef fish community composition across all families. With more than 200 fish species across all families, this is a tough ask for PCA so we will move to distance based ordinations.

- Group the data by transect, site, exposure, and species, and sum up the number of individuals (abundance). Turn the data into wide format and split it into meta and raw data as previously. 

```{r, echo = FALSE, results = FALSE}
# call the entire dataset
# group by transect, site, exposure, and species
# summarize the species abundances
lizard.sum <- lfish %>%
  group_by(transect, Site, Exposure, species) %>%
  summarize(abundance = sum(abundance))

# convert data to wide format
lizard.wide <- lizard.sum %>%
  spread(key = species, value = abundance, fill = 0)

# isolate metadata
lizard.meta <- lizard.wide[1:3]

# isolate numeric data
lizard.raw <- lizard.wide %>%
  ungroup() %>%
  select(-c(1:3))
```

- Create a distance matrix using the *vegdist* function and a Bray-Curtis dissimilarity metric. Then, run a PCoA using the *pcoa* function, choosing the cailliez correction for negative eigenvalues. Visualize the output using ggplot. 

```{r, echo = FALSE, results = FALSE}
# create a distance matrix with vegdist() from the vegan package
# use the Bray-Curtis dissimilarity metric, which is a dissimilarity measure designed to handle abundance data
lizard.dist <- vegdist(lizard.raw, method = "bray")

# run the PCoA using the cmdscale() function
lizard.pcoa <- pcoa(lizard.dist, correction = "cailliez")
lizard.pcoa

# extract the site scores and merge with metadata
lizard.scores <- as.data.frame(lizard.pcoa$vectors) %>%
  bind_cols(lizard.meta)

# create convex hulls grouped by site exposure based on site scores
lizard.hulls <- lizard.scores %>%
  group_by(Exposure) %>%
  slice(chull(Axis.1, Axis.2))

# plot PCoA with color and convex hulls as exposure
# you cannot add loadings to a PCoA because it only works as a distance matrix
lizard.plot.pcoa <- ggplot() +
  geom_point(data = lizard.scores, aes(x = Axis.1, y = Axis.2, color = Exposure), size = 2) +
  geom_polygon(data = lizard.hulls, aes(x = Axis.1, Axis.2, fill = Exposure), alpha = 0.5) +
  scale_fill_viridis_d() +
  scale_color_viridis_d() +
  theme_bw() +
  xlab("PCo1 (18.7%)") +
  ylab("PCo2 (12.6%)") +
  ggtitle("PCoA")
lizard.plot.pcoa
```


- run a non-metric multidimensional scaling ordination (NMDS) on the raw data using the *metaMDS* function, a Bray-Curtis dissimilarity metric, and two dimensions. Check the stress and plot the results using ggplot. Investigate your PCoA and NMDS side-by-side. 

```{r, echo = FALSE, results = FALSE}
# use metaMDS() to run the nMDS
# specify the data, the dimensions, the number of tries to give the nMDS to find a solution, and the distance metric
lizard.mds <- metaMDS(comm = lizard.raw, k = 2, trymax = 1000, distance = "bray")

# view results of nMDS
lizard.mds

# view stressplot
stressplot(lizard.mds)

# extract the site scores and merge with metadata
lizard.scores <- as.data.frame(lizard.mds$points) %>%
  bind_cols(lizard.meta)

# create convex hulls grouped by site exposure based on site scores
lizard.hulls <- lizard.scores %>%
  group_by(Exposure) %>%
  slice(chull(MDS1, MDS2))

# plot nMDS with color and convex hulls as exposure
lizard.plot.mds <- ggplot() +
  geom_point(data = lizard.scores, aes(x = MDS1, y = MDS2, color = Exposure), size = 2) +
  geom_polygon(data = lizard.hulls, aes(x = MDS1, MDS2, fill = Exposure), alpha = 0.5) +
  scale_fill_viridis_d() +
  scale_color_viridis_d() +
  theme_bw() +
  xlab("MDS1") +
  ylab("MDS2") +
  ggtitle("nMDS")
lizard.plot.mds

# view PcOA and nMDS together for comparison
# despite using the same distance metric (Bray-Curtis), the nMDS produces a cleaner, better separation among the transects than the PCoA.
pcoamds <- lizard.plot.pcoa / lizard.plot.mds
pcoamds
```

### Part IV

We will continue using the "fishcoms_lizardisland.csv" dataset, but this time, we will explore differences in fish communities between the two years, 2011 and 2015.

1) Calculate fish abundance per transect, site, and year for each species.
2) Put the dataset in wide format.
3) Split the metadata from the numeric data.

```{r, echo = FALSE, results = FALSE}
lizard.sum <- lfish %>%
  group_by(Site, transect, year, species) %>%
  summarize(abundance = sum(abundance))

lizard.wide <- lizard.sum %>%
  spread(key = species, value = abundance, fill = 0)

lizard.meta <- lizard.wide[1:3]

lizard.raw <- lizard.wide %>%
  ungroup() %>%
  select(-c(1:3))
```

1) Run a PCoA with Bray-Curtis dissimilarity.
2) Check eigenvalues. Add a correction if needed.
3) Plot the PCoA with different colors and polygons for the two years.

```{r, echo = FALSE, results = FALSE}
library(fishualize)
lizard.d.bray <- vegdist(lizard.raw, method = "bray")

lizard.pcoa.bray <-  pcoa(lizard.d.bray)

lizard.pcoa.bray <-  pcoa(lizard.d.bray, correction = "cailliez")

lizard.scores.b <- as.data.frame(lizard.pcoa.bray$vectors) %>%
  bind_cols(lizard.meta)

lizard.hulls.b <- lizard.scores.b %>%
  group_by(year) %>%
  slice(chull(Axis.1, Axis.2))

lizard.plot.b <- ggplot() +
  geom_point(data = lizard.scores.b, aes(x = Axis.1, y = Axis.2, color = as.factor(year)), size = 2) +
  geom_polygon(data = lizard.hulls.b, aes(x = Axis.1, Axis.2, fill = as.factor(year)), alpha = 0.5) +
  scale_fill_fish_d(option = "Stegastes_nigricans", end = 0.8) +
  scale_color_fish_d(option = "Stegastes_nigricans", end = 0.8) +
  theme_bw() +
  xlab("PCo1 (12.5%)") +
  ylab("PCo2 (8.4%)") +
  ggtitle("PCoA") +
  guides(color = guide_legend(title = "Year"), fill = "none")
lizard.plot.b
```


# d. 5: Solutions

- Start by installing and loading the *vegan* and the *ape* package into your environment, along with *dplyr*, *ggplot* and *patchwork* (or load the former two via the *tidyverse* package)

- Then, read in the fishcoms_lizardisland dataset, which is available above

### Part I

- Retain only species in the family Chaetodontidae (butterflyfishes) in the dataset, and summarize each species' abundance across sites, while also retaining information on each site's exposure level. Bring the data to wide format.

```{r, echo = TRUE, results = TRUE}
library(vegan)
library(tidyverse)
library(ape)
library(patchwork)
# re-familiarize with the Lizard Island fish community dataset
lfish <- read.csv(file =  "data/fishcoms_lizardisland.csv")
head(lfish)
lizard.sum <- lfish %>%
  filter(Family == "Chaetodontidae") %>%
  group_by(Site, Exposure, species) %>%
  summarize(abundance = sum(abundance))
lizard.sum

# spread into wide format and get rid of NAs
lizard.wide <- lizard.sum %>%
  spread(key = species, value = abundance, fill = 0)
lizard.wide
```

- Divide your dataset into metadata and abundance data, run a PCA on the Hellinger transformed data, and plot the data using ggplot. See how the data changes depending on whether you scale and center your data. 

```{r, echo = TRUE, results = TRUE}
# isolate metadata
lizard.meta <- lizard.wide[1:2]
lizard.meta

# isolate numeric data
lizard.raw <- lizard.wide %>%
  ungroup() %>%
  select(-c(1:2))
lizard.raw

# the decostand() function in vegan allows for transformations like Hellinger
lizard.hell <- decostand(lizard.raw, method = "hellinger")
head(lizard.hell)

# run a PCA and retain percent variation explained by PC1 and PC2
lizard.pca <- summary(rda(lizard.hell, scale = T))
lizard.pca

# extract the site scores and merge with metadata
lizard.scores <- as.data.frame(lizard.pca$sites) %>%
  bind_cols(lizard.meta)

# create convex hulls grouped by site exposure based on site scores
lizard.hulls <- lizard.scores %>%
  group_by(Exposure) %>%
  slice(chull(PC1, PC2))

# extract the loadings from species scores
lizard.loadings <- as.data.frame(lizard.pca$species) %>%
  mutate(species = rownames(.))

# plot the PCA results with color and convex hulls as exposure and species as loadings
lizard.plot2 <- ggplot() +
  geom_point(data = lizard.scores, aes(x = PC1, y = PC2, color = Exposure), size = 2) +
  geom_polygon(data = lizard.hulls, aes(x = PC1, PC2, fill = Exposure), alpha = 0.5) +
  geom_segment(data = lizard.loadings, aes(x = 0, xend = PC1*3,
                                           y = 0, yend = PC2*3), lwd = 0.1) +
  geom_text(data = lizard.loadings, aes(x = PC1*3, y = PC2*3, label = species), size = 3) +
  scale_fill_viridis_d() +
  scale_color_viridis_d() +
  theme_bw() +
  xlab("PC1 (21.8%)") +
  ylab("PC2 (17.8%)") +
  ggtitle("Hellinger transformed PCA, scale = T")
lizard.plot2

# run the same PCA, but with scale = F
# retain percent variation explained by PC1 and PC2
lizard.pca <- summary(rda(lizard.hell, scale = F))

# extract the site scores and merge with metadata
lizard.scores <- as.data.frame(lizard.pca$sites) %>%
  bind_cols(lizard.meta)

# create convex hulls grouped by site exposure based on site scores
lizard.hulls <- lizard.scores %>%
  group_by(Exposure) %>%
  slice(chull(PC1, PC2))

# extract the loadings from species scores
lizard.loadings <- as.data.frame(lizard.pca$species) %>%
  mutate(species = rownames(.))

# plot the PCA results with color and convex hulls as exposure and species as loadings
lizard.plot3 <- ggplot() +
  geom_point(data = lizard.scores, aes(x = PC1, y = PC2, color = Exposure), size = 2) +
  geom_polygon(data = lizard.hulls, aes(x = PC1, PC2, fill = Exposure), alpha = 0.5) +
  geom_segment(data = lizard.loadings, aes(x = 0, xend = PC1*2,
                                           y = 0, yend = PC2*2), lwd = 0.1) +
  geom_text(data = lizard.loadings, aes(x = PC1*2, y = PC2*2, label = species), size = 3) + # you can spread out loadings by multiplying values
  scale_fill_viridis_d() +
  scale_color_viridis_d() +
  theme_bw() +
  xlab("PC1 (30.3%)") +
  ylab("PC2 (20.5%)") +
  ggtitle("Hellinger transformed PCA, scale = F")
lizard.plot3
```

Part II 

- We now want to examine how the four regimes of wave action impact reef fish community composition across all families. With more than 200 fish species across all families, this is a tough ask for PCA so we will move to distance based ordinations.

- Group the data by transect, site, exposure, and species, and sum up the number of individuals (abundance). Turn the data into wide format and split it into meta and raw data as previously. 

```{r, echo = TRUE, results = TRUE}
# call the entire dataset
# group by transect, site, exposure, and species
# summarize the species abundances
lizard.sum <- lfish %>%
  group_by(transect, Site, Exposure, species) %>%
  summarize(abundance = sum(abundance))

# convert data to wide format
lizard.wide <- lizard.sum %>%
  spread(key = species, value = abundance, fill = 0)

# isolate metadata
lizard.meta <- lizard.wide[1:3]

# isolate numeric data
lizard.raw <- lizard.wide %>%
  ungroup() %>%
  select(-c(1:3))
```

- Create a distance matrix using the *vegdist* function and a Bray-Curtis dissimilarity metric. Then, run a PCoA using the *pcoa* function, choosing the cailliez correction for negative eigenvalues. Visualize the output using ggplot. 

```{r, echo = TRUE, results = TRUE}
# create a distance matrix with vegdist() from the vegan package
# use the Bray-Curtis dissimilarity metric, which is a dissimilarity measure designed to handle abundance data
lizard.dist <- vegdist(lizard.raw, method = "bray")

# run the PCoA using the cmdscale() function
lizard.pcoa <- pcoa(lizard.dist, correction = "cailliez")
lizard.pcoa

# extract the site scores and merge with metadata
lizard.scores <- as.data.frame(lizard.pcoa$vectors) %>%
  bind_cols(lizard.meta)

# create convex hulls grouped by site exposure based on site scores
lizard.hulls <- lizard.scores %>%
  group_by(Exposure) %>%
  slice(chull(Axis.1, Axis.2))

# plot PCoA with color and convex hulls as exposure
# you cannot add loadings to a PCoA because it only works as a distance matrix
lizard.plot.pcoa <- ggplot() +
  geom_point(data = lizard.scores, aes(x = Axis.1, y = Axis.2, color = Exposure), size = 2) +
  geom_polygon(data = lizard.hulls, aes(x = Axis.1, Axis.2, fill = Exposure), alpha = 0.5) +
  scale_fill_viridis_d() +
  scale_color_viridis_d() +
  theme_bw() +
  xlab("PCo1") +
  ylab("PCo2") +
  ggtitle("PCoA")
lizard.plot.pcoa
```


- run a non-metric multidimensional scaling ordination (NMDS) on the raw data using the *metaMDS* function, a Bray-Curtis dissimilarity metric, and two dimensions. Check the stress and plot the results using ggplot. Investigate your PCoA and NMDS side-by-side. 

```{r, echo = TRUE, results = TRUE}
# use metaMDS() to run the nMDS
# specify the data, the dimensions, the number of tries to give the nMDS to find a solution, and the distance metric
lizard.mds <- metaMDS(comm = lizard.raw, k = 2, trymax = 1000, distance = "bray")

# view results of nMDS
lizard.mds

# view stressplot
stressplot(lizard.mds)

# extract the site scores and merge with metadata
lizard.scores <- as.data.frame(lizard.mds$points) %>%
  bind_cols(lizard.meta)

# create convex hulls grouped by site exposure based on site scores
lizard.hulls <- lizard.scores %>%
  group_by(Exposure) %>%
  slice(chull(MDS1, MDS2))

# plot nMDS with color and convex hulls as exposure
lizard.plot.mds <- ggplot() +
  geom_point(data = lizard.scores, aes(x = MDS1, y = MDS2, color = Exposure), size = 2) +
  geom_polygon(data = lizard.hulls, aes(x = MDS1, MDS2, fill = Exposure), alpha = 0.5) +
  scale_fill_viridis_d() +
  scale_color_viridis_d() +
  theme_bw() +
  xlab("MDS1") +
  ylab("MDS2") +
  ggtitle("nMDS")
lizard.plot.mds

# view PcOA and nMDS together for comparison
# despite using the same distance metric (Bray-Curtis), the nMDS produces a cleaner, better separation among the transects than the PCoA.
pcoamds <- lizard.plot.pcoa / lizard.plot.mds
pcoamds
```

### Part IV

We will continue using the "fishcoms_lizardisland.csv" dataset, but this time, we will explore differences in fish communities between the two years, 2011 and 2015.

1) Calculate fish abundance per transect, site, and year for each species.
2) Put the dataset in wide format.
3) Split the metadata from the numeric data.

```{r, echo = TRUE}
lizard.sum <- lfish %>%
  group_by(Site, transect, year, species) %>%
  summarize(abundance = sum(abundance))

lizard.wide <- lizard.sum %>%
  spread(key = species, value = abundance, fill = 0)

lizard.meta <- lizard.wide[1:3]

lizard.raw <- lizard.wide %>%
  ungroup() %>%
  select(-c(1:3))
```

1) Run a PCoA with Bray-Curtis dissimilarity.
2) Check eigenvalues. Add a correction if needed.
3) Plot the PCoA with different colors and polygons for the two years.

```{r, echo = TRUE}
library(fishualize)
lizard.d.bray <- vegdist(lizard.raw, method = "bray")

lizard.pcoa.bray <-  pcoa(lizard.d.bray)

lizard.pcoa.bray <-  pcoa(lizard.d.bray, correction = "cailliez")

lizard.scores.b <- as.data.frame(lizard.pcoa.bray$vectors) %>%
  bind_cols(lizard.meta)

lizard.hulls.b <- lizard.scores.b %>%
  group_by(year) %>%
  slice(chull(Axis.1, Axis.2))

lizard.plot.b <- ggplot() +
  geom_point(data = lizard.scores.b, aes(x = Axis.1, y = Axis.2, color = as.factor(year)), size = 2) +
  geom_polygon(data = lizard.hulls.b, aes(x = Axis.1, Axis.2, fill = as.factor(year)), alpha = 0.5) +
  scale_fill_fish_d(option = "Stegastes_nigricans", end = 0.8) +
  scale_color_fish_d(option = "Stegastes_nigricans", end = 0.8) +
  theme_bw() +
  xlab("PCo1 (12.5%)") +
  ylab("PCo2 (8.4%)") +
  ggtitle("PCoA") +
  guides(color = guide_legend(title = "Year"), fill = "none")
lizard.plot.b
```
