---
title: "Session 6: Functional trait based analyses"
description: |
  This session introduces trait-based analyses as means to analyze communities under the consideration of ecological differences. 
author:
  - name: Simon J. Brandl, PhD
    url: https://www.fishandfunctions.com/
    affiliation: The University of Texas at Austin
date: "`r Sys.Date()`"
color1: "#0D2A47"
output:
  distill::distill_article:
    toc: true
    number_sections: true
    toc_depth: 4
    code_folding: false
    toc_float: true
    collapsed: false
    smooth_scroll: true
editor_options: 
  chunk_output_type: console
---


# a. 6: Exercises

- Start by loading the following packages (remember that you may have to install some of them): 

```{r, echo = TRUE}
library(tidyverse)
library(fishualize)
library(patchwork)
library(vegan)
library(FD)
library(patchwork)
library(ggformula)
```

```{r, echo = FALSE}
# traits <- read.csv(file = "data/traits_biomass_comb.csv") %>%
#   filter(Species != "sco.line") %>%
#   select(Species, Diet, Habitat, Territorial, Sizeclass)
# 
# sites <- read.csv(file = "data/traits_biomass_comb.csv") %>%
#   select(-Diet, -Habitat, -Territorial, -Sizeclass, - ID) %>%
#   pivot_longer(cols = 2:29, names_to = "site", values_to = "biomass") %>%
#   filter(biomass >0) %>%
#   pivot_wider(names_from = Species, values_from = biomass, values_fill = 0) %>%
#   mutate(year = case_when(grepl("2011", site) ~ 2011,
#                           TRUE ~ 2015))%>%
#   relocate(site, year)
# 
# write.csv(traits, file = "data/traits.csv", row.names = F)
# write.csv(sites, file = "data/sites.csv", row.names = F)
```

Data wrangling:

- We will first work with a reef fish dataset that was used for trait-based analyses in Brandl et al. 2016 (Ecosphere). The trait data are [here](data/traits.csv) and the site data, which contains biomass calculations for each community, are [here](data/reeffishcomms.csv), which we will call "sites". Load the data into your environment and then use the "column_to_rownames" function to turn the "Species" column (trait dataset) into your rownames. 

- For the sites dataset, remove the year column from the dataset and then use the "column_to_rowname" function again, this time for the "site" column. Then turn the sites data.frame into a matrix using the "as.matrix" function. 

- Finally, check that the rownames of the trait dataset match the column names of the sites dataset

```{r, echo = TRUE, results = TRUE}
traits = read.csv(file = "data/traits.csv")
sites = read.csv(file = "data/sites.csv")

traits1 <- traits %>%  
  mutate(Diet = as.factor(Diet),
         Habitat = as.factor(Habitat),
         Territorial = as.factor(Territorial),
         Sizeclass = as.factor(Sizeclass)) %>%
    column_to_rownames("Species")

sites1 <- sites %>%
  select(-year) %>%
  column_to_rownames("site") %>%
  as.matrix()

rownames(traits1) == colnames(sites1)
```


Running FD:

- To calculate the most common functional indices, we will use the dbFD function. Specify the traits dataset and the sites dataset as the two inputs, choose a correction for negative eigenvalues (usually cailliez or lingoes), and make sure to specify "print.pco = TRUE".

- Pull the number of species (nbsp), functional richness (FRic), functional evenness (FEve), and functional dispersion (FDis) values into a new dataset that also has your sites and the year (from the original sites dataset)

```{r, echo = TRUE, results = TRUE}
calc <- dbFD(x = traits1, a = sites1, corr = "lingoes", print.pco = T)


functional_metrics <- sites %>%
  select(site, year) %>%
  bind_cols(calc$nbsp, calc$FRic, calc$FEve, calc$FDis) %>%
  rename("Spric" = "...3",
    "FRichness" = "...4",
         "FEveness" = "...5",
         "FDispersion" = "...6")
```

Plotting: 

- Plot species richness against functional richness, fitting a line that suggests the best fit to the data

- Plot the four indices in 2011 and 2015 using boxplots, violins, or similar

- Plot the functional space across the two years, calling on the "x.axes" column in the output of the dbFD function

```{r, echo = TRUE, results = TRUE}
spric_fric <- ggplot(functional_metrics, aes(x = Spric, y = FRichness)) +
  geom_point(color = "sienna2") +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), color = "sienna2", fill = "sienna4") +
  theme_classic() +
  ylab("Functional richness") +
  xlab("Species richness")
spric_fric

functional_metrics_long <- functional_metrics %>%
  pivot_longer(cols = 3:6, names_to = "metric", values_to = "value")
  
diversity_plots <- ggplot(functional_metrics_long, aes(x = as.factor(year), y = value, fill = as.factor(year))) +
  geom_boxplot() +
  facet_wrap(.~metric, scales = "free") +
  scale_fill_manual(values = c("sienna2", "steelblue")) +
  theme_classic() +
  xlab("Year") +
  ylab("") +
  theme(legend.position = "none")
diversity_plots

### plot functional space

species.coords <- data.frame(calc$x.axes) %>%
  select(A1, A2) %>%
  mutate(species = rownames(.)) %>%
  mutate(A1 = round(A1, 4),
         A2 = round(A2, 4))

sites.coords <- sites %>%
  pivot_longer(cols = 3:275, names_to = "species", values_to = "biomass") %>%
  inner_join(species.coords) %>%
  filter(biomass > 0) %>%
  group_by(site, year, A1, A2) %>%
  summarize(biomass = sum(biomass))
  
functional.space <- ggplot(sites.coords, aes(x = A1, y = A2, fill = as.factor(year))) +
  geom_point(aes(size = biomass*10), shape = 21, alpha = 0.3) +
  facet_wrap(. ~ year) +
  scale_fill_manual(values = c("sienna2", "steelblue")) +
  theme_classic() +
  xlab("Axis1") +
  ylab("Axis2") +
  theme(legend.position = "none")
functional.space
```


